# Predation and Fleeing Behavior
```{r, include=FALSE}
current_chapter <- "15-Predation.Rmd"
```
```{r, child="_download_button.Rmd"}
```

## Overview

This module simulates predator–prey interactions between striped bass (predators) and alewife (prey) during migratory and staging periods, representing encounters as localized behavioral interactions embedded within movement, schooling, and energetic state. Predators visually detect nearby prey based on proximity, visual range, gape limitation, and stomach capacity, initiate pursuit when energetically feasible, and consume prey within capture range. Prey respond through directional fleeing and short evasive movements, with escape responses propagating across nearby individuals through a patch-level alarm mechanism. The model tracks energetic costs of pursuit and escape, prey consumption events, and resulting spatial patterns of predation risk.

The interaction structure is adapted from the NetLogo schooling and predation framework of [@dobon_netlogo_2018], extended here by linking chase, escape, and capture outcomes to hydrodynamic movement, energetic cost, and contaminant transfer. Predator–prey interactions are inherently spatial and emerge from movement and local overlap among individuals, where encounter rates are governed by proximity and relative motion through the environment [@northfield_spatial_2017]. These interactions follow established predator–prey theory in which consumption arises from encounter processes and functional responses linking prey availability to predator intake [@diz-pita_predatorprey_2021], allowing predation risk, prey displacement, predator foraging success, and contaminant uptake to emerge dynamically rather than being imposed as fixed mortality or feeding rates.

## Purpose and Patterns

This submodel captures behaviorally realistic predator–prey dynamics relevant to estuarine migration and staging.

-   Prey initiate fleeing when striped bass are within a defined visual neighborhood and prey have sufficient energy to respond.

-   Predators initiate pursuit when prey are visible and the predator is not constrained by energetic state or stomach capacity.

-   Fleeing behavior changes prey movement mode and overrides normal migration movement.

-   Predation events occur when predators and prey co locate within a capture neighborhood and prey meet gape constraints.

-   Predation risk hotspots emerge from the overlap of predator movement, prey distribution, and hydrodynamic movement constraints.

## Entities, State Variables, and Scales

### Spatial and Temporal Scales

**Spatial Unit:** Patch (3 m x 3 m resolution)\
**Temporal Unit:** 5 minute time steps (tick)

### Patch Variables

| Variable Name | Definition |
|------------------------------------|------------------------------------|
| `patch-terrain` | Patch classification used to restrict movement to water patches. |
| `prey-eaten-in-patch` | Count of prey consumed in a patch. |
| `visits-by-alewife` | Cumulative count of prey patch visits, used for exposure and space use summaries. |
| `ticks-spent-alewife` | Cumulative time spent by prey on a patch. |

### Prey: Agent Variables

| Variable Name | Definition |
|------------------------------------|------------------------------------|
| `fleeing?` | Boolean indicator of whether the prey is actively fleeing. |
| `energy` | Energetic state used to determine whether prey can flee and to regulate behavior switching. |
| `speed` | Current movement speed used in the swim procedure. |
| `max-speed` | Maximum sustained swim speed. |
| `vision-distance` | Detection or interaction distance used in predator encounter checks. |
| `planned-path` | Stored movement path used during fleeing movement. |
| `trail` | List of visited patches retained for optional visualization. |
| `previous-patch`, `previous-x`, `previous-y` | Tracking variables supporting movement diagnostics and behavior switching. |

### Predator: Agent Variables

| Variable Name | Definition |
|------------------------------------|------------------------------------|
| `vision-distance` | Radius within which prey are considered visible to the predator. |
| `prey-in-vision` | Agentset of visible prey within the predator’s vision radius. |
| `hunting?` | Boolean indicator of whether the predator is actively pursuing prey. |
| `bursting?` | Boolean indicator of whether the predator is in burst speed mode. |
| `speed` | Predator movement speed, which can increase during bursting. |
| `max-speed` | Maximum predator speed used during chase. |
| `stomach-contents` | Current stomach contents used to limit feeding and trigger foraging and lipid use when empty. |
| `stomach-capacity` | Maximum stomach capacity. |
| `gape-size` | Predator gape size used as a constraint on edible prey. |
| `daytime-prey-eaten` | Daily counter of prey eaten. |
| `total-prey-eaten` | Cumulative prey eaten (tracked via counters). |
| `prey-eaten-in-patch` | Patch level counter incremented during predation. |
| `E-swim` | Swimming energy cost computed during movement. |
| `difficulty-factor` | Hydrodynamic movement difficulty used to scale energetic cost of movement. |
| `swim-efficiency` | Efficiency factor used to scale energetic cost of movement. |

## Process Overview and Scheduling

**Prey Behavior (per agent)**

1.  During `swim`, if `fleeing?` is false, prey execute normal migration
    movement and pay swimming energetic costs.

2.  If `fleeing?` is true, prey execute `flee-stripedbass`, which
    overrides normal migration movement and moves prey through a short
    random evasive path over water patches.

3.  Prey stop fleeing when no striped bass are detected within the
    configured encounter range or when prey energy is depleted.

**Predator Behavior (per agent)**

1.  During active movement periods, predators define
    $\mathcal{P}_{vision}$ as prey within $A_{vision}$.

2.  If stomach contents are empty, predators enter foraging mode and may
    activate lipid loss depending on the broader energy logic.

3.  If any prey are in vision, predators set `hunting?` to true, chase
    prey, and attempt capture.

4.  Predators adjust speed during pursuit, trigger prey fleeing
    behaviors, and consume prey when capture criteria are met.

5.  Predation updates predator stomach contents, prey counters, and
    patch level consumption counters.

## Design Concepts

**Basic Principles:** Predator–prey interactions are adapted from the
[@dobon_netlogo_2018] NetLogo schooling and predation framework. The
core structure of local visual detection, directional pursuit, prey
fleeing, and stochastic capture is retained from that implementation. In
P-MEM, these interaction rules are extended by coupling chase and escape
behavior to energetic constraints, gape limitation, stomach capacity,
and contaminant transfer. Thus, the underlying encounter mechanics
broadly follow @dobon_netlogo_2018, while cost-routing, physiological
scaling and contaminant dynamics represent model extensions.

**Emergence:** Spatial predation patterns emerge from coupled movement
processes, including prey migration routes, predator search behavior,
and the temporary displacement of prey caused by fleeing responses.

**Adaptation:** Prey adaptively switch from migration movement to
fleeing movement when predators are nearby. Predators adaptively switch
between hunting and wandering depending on prey availability and stomach
state.

**Objectives:** Predators maximize energetic intake by capturing prey
when available while managing energetic costs of pursuit. Prey minimize
predation risk by fleeing when threatened, balanced against energetic
limitations.

**Sensing:** Predators and prey sense one another through neighborhood
or cone based proximity checks using $A_{vision}$. Prey fleeing is
triggered by local predator presence.

**Stochasticity:** Stochasticity arises through random turning and patch
selection during fleeing movement and through prey selection during
capture attempts.

**Collectives:** Collective dynamics are represented indirectly through
the simultaneous triggering of fleeing across multiple prey within the
predator’s encounter neighborhood.

**Observation:** Key outputs include prey consumption rates, prey and
predator energetic trajectories, and time periods of predation, and the
spatial distribution of predation events ($P_{eaten,patch}$).

## Initialization

| Variable | Initialized Value | Justification |
|:----------------------:|:----------------------:|:----------------------:|
| **energy** | 100 | Prey begin fully energized to allow immediate fleeing or normal movement |
| **fleeing?** | false | Prey are not actively escaping at the start |
| **prey-alarmed?** | false | Alarm cue inactive at start of simulation |
| **daytime-prey-eaten** | 0 | daily counter of prey eaten starts at zero |
| **time-since-full** | 0 | predator begins simulation hungry |
| **gape-limit** | function of size | Large predators have wider gape limits |
| **handling-effort** | calculated per encounter | Depends on prey traits and local patch conditions |
| **reaction-time** | size-, age-, and SPM-based | Prey and predator values dynamically computed each tick |
| **flee-ability** | based on age | Represents individual escape competence across age classes |
| **predation-ability** | based on age | Mid-age predators have fastest and most accurate prey responses |
| $swim_{max}$ | $1.5-3 \frac{body lengths}{sec}$ | Typical value for sustained swimming speed in small pelagic fish (refer to Videler, 1993). |

## Submodels

### Prey Fleeing Trigger and Directional Response

Prey enter a fleeing state when a striped bass is detected within a
cone-based encounter neighborhood and prey energy is positive:

$$
 \exists \; predator \in \mathcal{C}_{prey}(r,\theta)\;\; \land \;\; E_{agent} > 0 \;\;\Rightarrow\;\; fleeing? = \text{true}  \hspace{2cm} \text{(15.1)} 
$$ Where:

-   $\mathcal{C}_{prey}(r,\theta)$ is the prey encounter cone with
    radius $r = A_{vision}$ (patch units) and cone angle $\theta$
    (implemented as `in-cone A_{vision} 180`) (degrees).
-   $E_{agent}$ is the total internal energy available to the agent
    (patch units).

Prey exit fleeing when no striped bass are detected within the encounter
neighborhood or when energy is depleted:

$$
 \neg \exists \; predator \in \mathcal{C}_{prey}(r,\theta)\;\; \lor \;\; E_{agent} \le 0 \;\;\Rightarrow\;\; fleeing? = \text{false}  \hspace{2cm} \text{(15.2)} 
$$

Where:

-   $\mathcal{C}_{prey}(r,\theta)$ is the prey encounter cone with
    radius $r = A_{vision}$ (patch units) and cone angle $\theta$
    (implemented as `in-cone A_{vision} 180`) (degrees).
-   $E_{agent}$ is the total internal energy available to the agent
    (energy units).

Directional escape is implemented by assigning heading based on predator
relative position, corresponding to the `scare-prey` control flow:

$$
 \psi_{t+1} =
\begin{cases}
\psi_t + 90^\circ  & \text{(scare-right)}\\
\psi_t - 90^\circ  & \text{(scare-left)}\\
\psi_t + 180^\circ & \text{(scare-down)}\\
\psi_t + 0^\circ   & \text{(scare-up)}
\end{cases}  \hspace{2cm} \text{(15.3)} 
$$

Where:

-   $\mathcal{C}_{prey}(r,\theta)$ is the prey encounter cone with
    radius $r = A_{vision}$ (patch units) and cone angle $\theta$
    (implemented as `in-cone A_{vision} 180`) (degrees).
-   $\psi$ is the agent heading (degrees).

### Prey Fleeing Movement and Energetic Cost

When fleeing, prey move at maximum speed:

$$
 A_{speed} = speed_{max}  \hspace{2cm} \text{(15.4)} 
$$

Where:

-   $A_{speed}$ is the actual swimming speed used for movement (m s⁻¹).
-   $speed_{max}$ is the maximum achievable swimming speed (m s⁻¹).

Travel distance per tick is computed from current speed (with the
model’s internal conversion):

$$
 D_{travel} = \frac{A_{speed}}{3}  \hspace{2cm} \text{(15.5)} 
$$

Where:

-   $D_{travel}$ is the distance traveled in the current tick (patch
    units), where 1 patch = 3 m.
-   $A_{speed}$ is the actual swimming speed used for movement (m
    tick⁻¹).

A small random turn is applied:

$$
 \psi_{t+1} = \psi_t + \varepsilon,\;\;\; \varepsilon \sim U(-15^\circ, 15^\circ)  \hspace{2cm} \text{(15.6)} 
$$

Where:

-   $\psi$ is the agent heading (degrees).
-   $\varepsilon$ is a random turning increment (degrees).

A target patch is selected from reachable water patches within distance
$D_{travel}$:

$$
 p^* \sim \text{Uniform}\left(\{p \in \mathcal{N}(d_t) : patch\text{-}terrain(p)=\text{"water"}\}\right)  \hspace{2cm} \text{(15.7)} 
$$

Where:

-   $p^*$ is the selected target patch (patch location).
-   $p$ is the patch occupied by the prey agent (patch location.
-   $\mathcal{N}(D_{travel})$ is the set of patches within radius
    $D_{travel}$.
-   $patch\text{-}terrain(p)$ returns the terrain type of patch $p$.
-   $\text{Uniform}(\cdot)$ indicates uniform random selection.

Prey traverse a path from the current patch $p$ to $p^*$ (implementation
uses `build-path-to-prey`) and move along up to

$$
 n_t = \lceil D_{trave} \rceil  \hspace{2cm} \text{(15.8)} 
$$

Where:

-   $n_t$ is the number of patch steps taken within the tick (patches).
-   $\lceil \cdot \rceil$ denotes the ceiling operator.

patch steps within the tick. Patch visitation and time spent are
accumulated:

$$
 visits\text{-}by\text{-}alewife(p) \leftarrow visits\text{-}by\text{-}alewife(p) + 1  \hspace{2cm} \text{(15.9)} 
$$

$$
 ticks\text{-}spent\text{-}alewife(p) \leftarrow ticks\text{-}spent\text{-}alewife(p) + 1  \hspace{2cm} \text{(15.10)} 
$$

Where:

-   $p$ is the patch occupied by the prey agent (patch location).
-   $visits\text{-}by\text{-}alewife(p)$ records cumulative visitation
    (visits).
-   $ticks\text{-}spent\text{-}alewife(p)$ records cumulative residence
    time (ticks).

Swimming energetic cost is applied through the shared swimming energy
routine. First compute an energy multiplier from hydrodynamic difficulty
and swimming efficiency:

$$
 M_{swim} = D_f \cdot \left(\frac{1}{swim_{Eff}}\right)  \hspace{2cm} \text{(15.11)} 
$$

Where:

-   $D_f$ is mapped swimming difficulty index (dimensionless).
-   $swim_{Eff}$ is the efficiency multiplier applied to swimming
    energetics (%).
-   $M_{swim}$ is the swimming energy multiplier (dimensionless).

Then compute swimming energy cost (exponential scaling):

$$
 E_{swim} = swim_{base} \cdot m_{swim}^{\beta}  \hspace{2cm} \text{(15.12)} 
$$

Where:

-   $E_{swim}$ is the energy cost of swimming for the tick (energy
    units).
-   $swim_{base}$ is the baseline energetic cost of swimming ((energy
    units) patch⁻¹).
-   $M_{swim}$ is the swimming energy multiplier (dimensionless).
-   $\beta$ is the exponent controlling swimming cost scaling
    (dimensionless.

with $\beta = 0.75$. Energy is updated and constrained to remain
non-negative:

$$
 energy_{t+1} = \max(0,\; energy_t - E_{swim,t})  \hspace{2cm} \text{(15.13)} 
$$

Where:

-   $D_f$ is the mapped swimming difficulty index (dimensionless).
-   $swim_{Eff}$ is the efficiency multiplier applied to swimming
    energetics (%).
-   $M_{swim}$ is the swimming energy multiplier (dimensionless).
-   $swim_{base}$ is the baseline energetic cost of swimming ((energy
    units) patch⁻¹).
-   $\beta$ is the exponent controlling swimming cost scaling
    (dimensionless).
-   $E_{agent}$ is the total internal energy available to the agent
    (energy units).

### Predator Prey Detection and Chase

Predators define visible prey as prey within a radius equal to
$A_{vision}$:

$$
 \mathcal{P}_{vision} = \{i \in \text{prey} : \|x_i - x_{pred}\| \le A_{vision} \;\land\; patchtype(i) \ne \text{"sea"}\}  \hspace{2cm} \text{(15.14)} 
$$

Where:

-   $\mathcal{P}_{vision}$ is the set of prey visible to the predator.
-   $i$ indexes individual prey agents.
-   $x_i$ is the spatial position vector of prey $i$ (patch
    coordinates).
-   $x_{pred}$ is the spatial position vector of the predato (patch
    coordinates)r.
-   $\|\cdot\|$ denotes Euclidean distance (patch units).
-   $A_{vision}$ is the predator's visual detection radius (patch
    units).
-   $patchtype(i)$ returns the terrain type occupied by prey $i$.

Predators enter hunting mode when any prey are visible:

$$
 |\mathcal{P}_{vision}| > 0 \Rightarrow hunting? = \text{true}  \hspace{2cm} \text{(15.15)} 
$$

Where:

-   $\mathcal{P}_{vision}$ is the set of prey visible to the predator.
-   $hunting?$ is the indicator that predator is hunting.

If hunting, the predator targets the nearest visible prey:

$$
 i^* = \arg\min_{i \in \mathcal{P}_{vision}} \; \|x_i - x_{pred}\|  \hspace{2cm} \text{(15.16)} 
$$

Where:

-   $i^*$ is the selected target prey.
-   $\arg\min$ returns the prey minimizing distance to the predator
    (patch units).
-   $x_i$ is the spatial position vector of prey $i$ (patch
    coordinates).
-   $x_{pred}$ is the spatial position vector of the predator (patch
    coordinates).

Predator movement toward prey is computed using a step length limited by
travel distance per tick and prey distance:

$$
 D_{travel} = \frac{A_{speed}}{3}  \hspace{2cm} \text{(15.17)} 
$$ Where:

-   $D_{travel}$ is the distance traveled in the current tick (patch
    units), where 1 patch = 3 m.
-   $A_{speed}$ is the actual swimming speed used for movement (m
    tick⁻¹).

$$
 step_t = \min(D_{travel},\; \|x_{i^*} - x_{pred}\|)  \hspace{2cm} \text{(15.18)} 
$$

Where:

-   $step_{t}$ is the realized movement distance toward the selected
    target during the current tick.
-   $D_{travel}$ is the distance traveled in the current tick (patch
    units), where 1 patch = 3 m.
-   $x_i$ is the spatial position vector of prey $i$ (patch
    coordinates).
-   $x_{pred}$ is the spatial position vector of the predator (patch
    coordinates).

$$
 x_{pred+1} = x_{pred} + step_t \cdot \hat{u}(x_{i^*} - x_{pred})  \hspace{2cm} \text{(15.19)} 
$$

Where:

-   $x_{pred}$ is the spatial position vector of the predator (patch
    coordinates).
-   $step_{t}$ is the realized movement distance toward the selected
    target during the current tick (patch units).
-   $\hat{u}$ is the unit direction vector toward the target prey
    (dimensionless).
-   $x_{i^*}$ is the spatial position vector of prey $i$ (patch
    coordinates).

Swimming energetic cost during chase is applied using the same shared
routine:

$$
 E_{swim,t} = swim_{base} \cdot \left(D_f \cdot \frac{1}{swim_{Eff}}\right)^{\beta}  \hspace{2cm} \text{(15.20)} 
$$

Where

-   $E_{swim}$ is the energy cost of swimming for the tick (energy
    units).
-   $D_f$ is the mapped swimming difficulty index (dimensionless).
-   $swim_{Eff}$ is the efficiency multiplier applied to swimming
    energetics (%).
-   $swim_{base}$ is the baseline energetic cost of swimming ((energy
    units) patch⁻¹).
-   $\beta$ is the exponent controlling swimming cost scaling
    (dimensionless).

$$
 E_{agent+1} = \max(0,\; E_{agent} - E_{swim,t})  \hspace{2cm} \text{(15.21)} 
$$

Where:

-   $\mathcal{P}_{vision}$ is the set of visible prey.
-   $x_i*$ is the spatial position vector of prey $i$ (patch
    coordinates).
-   $\hat{u}$ is the unit direction vector toward prey (patch
    coordinates).
-   $step_{t}$ is the realized movement distance toward the selected
    target during the current tick (patch units).
-   $E_{agent}$ is the total internal energy available to the agent
    (energy units).
-   $E_{swim}$ is the energy cost of swimming for the tick (energy
    units).

### Predator Burst Speed Adjustment

Predators define visible prey as prey within a radius equal to
$A_{vision}$:

$$
 \mathcal{P}_{vision} = \{i \in \text{prey} : \|x_{i^*} - x_{pred}\| \le A_{vision} \;\land\; patchtype(i) \ne \text{"sea"}\}  \hspace{2cm} \text{(15.22)} 
$$

Where:

-   $A_{vision}$ is the visual detection radius of agent (patch units).
-   $\mathcal{P}_{vision}$ is the set of visible prey.
-   $x_{i^*}$ is the spatial position vector of prey $i$ (patch
    coordinates).
-   $x_{pred}$ is the spatial position vector of the predator (patch
    coordinates).
-   $patchtype$ is the location classifier of the patch.

Predators enter hunting mode when any prey are visible:

$$
 |\mathcal{P}_{vision}| > 0 \Rightarrow hunting? = \text{true}  \hspace{2cm} \text{(15.23)} 
$$

Where:

-   $\mathcal{P}_{vision}$ is the set of visible prey.
-   $hunting?$ is the indicator that predator is hunting.

If hunting, the predator targets the nearest visible prey:

$$
 i^* = \arg\min_{i \in \mathcal{P}_{vision}} \; \|x_{i^*} - x_{pred}\|  \hspace{2cm} \text{(15.24)} 
$$

Where:

-   $i^*$ is the selected target prey.
-   $\mathcal{P}_{vision}$ is the set of visible prey.
-   $x_{i^*}$ is the spatial position vector of prey $i$ (patch
    coordinates).
-   $x_{pred}$ is the spatial position vector of the predator (patch
    coordinates).

Predator movement toward prey is computed using a step length limited by
travel distance per tick and prey distance:

$$
 D_{travel} = \frac{A_{speed}}{3}  \hspace{2cm} \text{(15.25)} 
$$

Where:

-   $D_{travel}$ is the distance traveled in the current tick (patch
    units), where 1 patch = 3 m.
-   $A_{speed}$ is the actual swimming speed used for movement (m
    tick⁻¹).

$$
 step_t = \min(D_{travel},\; \|x_{i^*} - x_{pred}\|)  \hspace{2cm} \text{(15.26)} 
$$

Where:

-   $step_{t}$ is the realized movement distance toward the selected
    target during the current tick.
-   $D_{travel}$ is the distance traveled in the current tick (patch
    units), where 1 patch = 3 m.
-   $x_{i^*}$ is the spatial position vector of prey $i$ (patch
    coordinates).
-   $x_{pred}$ is the spatial position vector of the predator (patch
    coordinates).

$$
 x_{pred+1} = x_{pred} + step_t \cdot \hat{u}(x_{i^*} - x_{pred})  \hspace{2cm} \text{(15.27)} 
$$

Where:

-   $x_{pred}$ is the spatial position vector of the predator (patch
    coordinates).
-   $step_{t}$ is the realized movement distance toward the selected
    target during the current tick (patch units).
-   $\hat{u}$ is the unit direction vector toward prey.
-   $x_{i^*}$ is the spatial position vector of prey $i$ (patch
    coordinates).

Swimming energetic cost during chase is applied using the same shared
routine:

$$
 E_{swim} = swim_{base} \cdot \left(D_f \cdot \frac{1}{swim_{Eff}}\right)^{\beta}  \hspace{2cm} \text{(15.28)} 
$$

Where:

-   $E_{swim}$ is the energy cost of swimming for the tick (energy
    units).
-   $D_f$ is the mapped swimming difficulty index (dimensionless).
-   $swim_{Eff}$ is the efficiency multiplier applied to swimming
    energetics (%).
-   $swim_{base}$ is the baseline energetic cost of swimming ((energy
    units) patch⁻¹).
-   $\beta$ is the exponent controlling swimming cost scaling.

$$
 E_{agent+1} = \max(0,\; E_{agent} - E_{swim})  \hspace{2cm} \text{(15.29)} 
$$

Where:

-   $E_{agent}$ is the total internal energy available to the agent
    (energy units).
-   $E_{gain}$ is the energy gained from digested biomass (energy
    units).

### Predation and Consumption

Predators attempt capture only if stomach contents are below Predators can successfully capture only if stomach contents are below
capacity:

$$
 M_{stomach} < stomach_{max}  \hspace{2cm} \text{(15.30)} 
$$

Where:

-   $M_{stomach}$ is the biomass currently stored in the stomach (g).
-   $stomach_{max}$ is the maximum stomach capacity (g).

Candidate prey include prey on the same patch and neighboring patches
(local capture neighborhood):

$$
 \mathcal{P}_{near} = \{i \in \text{prey} : i \in \{here\} \cup \{neighbors\}\}  \hspace{2cm} \text{(15.31)} 
$$

Where:

-   $\mathcal{P}_{near}$ is a set of prey individuals located on the
    predator’s current patch.
-   $i$ is the index representing an individual prey agent.

Edible prey satisfy the gape constraint:

$$
 \mathcal{P}_{edible} = \{i \in \mathcal{P}_{near} : body\text{-}depth_i \le gape\text{-}size_{pred}\}  \hspace{2cm} \text{(15.32)} 
$$

Where:

-   $\mathcal{P}_{edible}$ is a subset of nearby prey that satisfy the
    predator’s gape limitation and are physically consumable.
-   $\mathcal{P}_{near}$ is a set of prey individuals located on the
    predator’s current patch.
-   $body\text{-}depth_i$ is the body depth of the selected prey
    individual (mm).
-   $gape\text{-}size_{pred}$ is the maximum ingestible prey depth for
    predator (mm).
-   $i$ is the index representing an individual prey agent.

If any edible prey exist, one prey is selected:

$$
 i^* \sim \text{Uniform}(\mathcal{P}_{edible})  \hspace{2cm} \text{(15.33)} 
$$

Where:

-   $i^*$ is the selected target prey.
-   $\mathcal{P}_{edible}$ is a subset of nearby prey that satisfy the
    predator’s gape limitation and are physically consumable.

Digestible meal mass is assumed to be 75% of prey wet weight:

$$
 M_{meal} = 0.75 \cdot A_{mass,i^*}  \hspace{2cm} \text{(15.34)} 
$$

Where:

-   $M_{meal}$ is the digestible biomass of captured prey (g).
-   $A_{mass,i^*}$ is the agent body mass of the selected target prey
    (g).

Available stomach space is:

$$
 S_{stomach} = stomach_{max} - M_{stomach}  \hspace{2cm} \text{(15.35)} 
$$

Where:

-   $S_{stomach}$ is the available stomach capacity (g).
-   $stomach_{max}$ is the maximum stomach capacity (g).
-   $M_{stomach}$ is the biomass currently stored in the stomach (g).

Intake is limited by available space:

$$
 M_{intake} = \min(M_{meal},\; S_{stomach})  \hspace{2cm} \text{(15.36)} 
$$

Where:

-   $M_{intake}$ is the biomass actually ingested by the agent (g).
-   $M_{meal}$ is the digestible biomass of captured prey (g).
-   $S_{stomach}$ is the available stomach capacity (g).

Stomach contents are updated:

$$
 M_{stomach+1} = M_{stomach} + M_{intake}  \hspace{2cm} \text{(15.37)} 
$$

Where:

-   $S_{stomach}$ is the available stomach capacity (g).
-   $M_{intake}$ is the biomass actually ingested by the agent (g).

Prey-derived contaminant loads are added to stomach contaminant pools:

$$
 stomach\text{-}contents\text{-}MeHg_{t+1} = stomach\text{-}contents\text{-}MeHg_t + MeHg_{i^*}  \hspace{2cm} \text{(15.38)} 
$$

Where:

-   $S_{stomach}$ is the available stomach capacity (g).
-   $MeHg_{i^*}$ methylmercury content of target prey (ng).

Consumption counters are incremented:

$$
 \mathcal{P}_{eaten+1} = \mathcal{P}_{eaten} + 1  \hspace{2cm} \text{(15.39)} 
$$

Where:

-   $\mathcal{P}_{eaten}$ is the total prey consumed by a predator.

$$
 \mathcal{P}_{eaten, patch+1} = \mathcal{P}_{eaten, patch} + 1  \hspace{2cm} \text{(15.40)} 
$$

Where:

-   $\mathcal{P}_{eaten, patch}$ is the amount of prey consumed on a
    given patch.

Selected prey is removed from the simulation:

$$
 alive(i^*) \leftarrow 0  \hspace{2cm} \text{(15.41)} 
$$

Where:

-   $alive(i^*)$ is the Boolean indicating whether selected prey is
    alive (0,1).

### Interaction With Migration Movement and Swimming Energy

Movement and swimming energetic cost are handled within the shared
movement controller `swim`. If prey are not fleeing, migration movement
is executed through STST procedures and swimming energy is charged. If
prey are fleeing, fleeing movement is executed and swimming energy is
charged. Therefore, no additional movement energy term is added within
predation or fleeing procedures:

$$
 E_{move} = E_{swim} \quad \text{applied once per tick via } calculate\text{-}swim\text{-}energy  \hspace{2cm} \text{(15.42)} 
$$

Where:

-   $E_{move}$ is the energy cost to move toward prey (energy units).

-   $E_{swim}$ is the energy cost of swimming for the tick (energy
    units).
    
## Netlogo Implementation

### Flee Predators

```{netlogo}
to flee-stripedbass
      if speed <= max-speed [
        set speed max-speed
          ;; how far can the fish travel this tick
      let travel-distance (speed / 3)
      let cur-loc patch-here
      
      ;; random small turn
      rt (random 30) - 15
      
      ;; patches in a cone in front, out to vision-distance
      let forward-patches patches in-radius travel-distance
      
      ;; keep only water patches that are within travel-distance
      let patch-options forward-patches with [
        patch-terrain = "water"
      ]
      
      ;; if no patch available → don't move
      if not any? patch-options [
        set planned-path [] 
        set patch-options neighbors with [patch-terrain = "water"]
      ]
      
      ;; pick a random forward patch
      let target-patch one-of patch-options
      
      ;; Determine shortest direct path (1-step or multi-step)
      let path build-path-to-prey cur-loc target-patch
      
      ;; remove the starting patch if it equals current position
      if first path = cur-loc [
        set path but-first path
      ]
      
      set planned-path path
      
      ;; If path exists
      if not empty? path [
        
        ;; How many patches to move this tick
        let steps-to-move ceiling travel-distance
        
        ;; Extract only the patches we will actually traverse
        let move-patches sublist path 0 (min list steps-to-move length path)
        ;; 6. If we moved, compute difficulty of the final patch
        let next-step last move-patches
        calculate-chase-difficulty next-step
        
        ;; Store starting patch
        let old-patch patch-here
        
        ;; Move along path step-by-step
        foreach move-patches [ p ->
          
          ;; -------- VISIT COUNTING --------
          if p != old-patch [
            ask p [
              set visits-by-alewife (visits-by-alewife + 1)
            ]
            set old-patch p
          ]
          
          ;; -------- MOVE --------
          face p
          move-to p
          
          ;; -------- TIME SPENT --------
          ask p [
            set ticks-spent-alewife (ticks-spent-alewife + 1)
          ]
        ]
        
        if not member? patch-here trail [
          set trail lput patch-here trail
        ]
      ]
      ]
      ;set rest-time alewife-rest-time
      ; set color green
end
```

### Scare Prey

```{netlogo}
to scare-prey ;; predator procedure
if any? alewives in-radius vision-distance [
  scare-right  ;; prey on right side will flee right
  rt 90
  scare-left  ;; prey on left side will flee left
  left 90
  scare-down ;; prey underneath will flee down
  right 180
  scare-up ;; prey underneath will flee down
  right 0
  ]
end

to scare-right  ;; prey on right side will flee right
 ask alewives in-cone vision-distance 180 [
    if any? stripedbass in-cone vision-distance 270 and energy > 0 [
      set fleeing? true
      right 90
    ]
    if not any? stripedbass in-cone vision-distance 270 or energy <= 0 [
      set fleeing? false
    ]

  ]
end

to scare-left ;; prey on the left side will flee left
  ask alewives in-cone vision-distance 180 [
    if any? stripedbass in-cone vision-distance 270 and energy > 0 [
      set fleeing? true
      left 90
    ]
    if not any? stripedbass in-cone vision-distance 270 or energy <= 0 [
      set fleeing? false
    ]
  ]
end

to scare-up ;; prey on left side will flee left
   ask alewives in-cone vision-distance 180 [
    if any? stripedbass in-cone vision-distance 270 and energy > 0 [
      set fleeing? true
      right 0
    ]
    if not any? stripedbass in-cone vision-distance 270 or energy <= 0 [
      set fleeing? false
    ]
  ]
end

to scare-down ;; prey on left side will flee left
   ask alewives in-cone vision-distance 180 [
    if any? stripedbass in-cone vision-distance 270 and energy > 0 [
      set fleeing? true
      right 180
    ]
    if not any? stripedbass in-cone vision-distance 270 or energy <= 0 [
      set fleeing? false
    ]
  ]
end
```

### Chase Nearest Prey

```{netlogo}
to chase-nearest-alewife ;; predator procedure
    let travel-distance (speed / 3)
    let nearest-alewife-prey min-one-of prey-in-vision [distance myself]
    if nearest-alewife-prey != nobody [
      face nearest-alewife-prey
      let d distance nearest-alewife-prey
      let step min list travel-distance d  ;; clamp
      fd step
    ]
  decide-chase-direction ; dynamic swimming difficulty
  calculate-swim-energy ; based on travel direction
end

to chase-nearest-alewife_old  ;; predator procedure
  ;; 1. Find prey and decide direction
  find-target-prey
  decide-chase-direction
  
  ;; 2. Calculate travel distance in patch units
  let travel-distance (speed / 3)

  ;; Default to no path
  set planned-path []
  
  if target-prey != nobody and patch-here = [patch-here] of target-prey [
  stop ;; no futher action needed
]
  ;; 3. Build a path only if prey exists
  if target-prey != nobody [
    
    let prey-loc [patch-here] of target-prey
    let cur-loc patch-here
    
    ;; Build cost map toward prey
    calculate-cost-to-prey prey-loc
    
    ;; Raw path from current patch to prey patch
    let raw-path build-path-to-prey cur-loc prey-loc

    ;; If path is non empty, drop the starting patch if it is the same as current location
    if not empty? raw-path [
      let first-step first raw-path
      let clean-path raw-path
      
      ;; If first element is current patch, skip it
      if first-step = patch-here [
        set clean-path but-first raw-path
      ]
      
      set planned-path clean-path
    ]
  ]

  ;; If there is no usable path, bail out for this tick
  if empty? planned-path [
    ;; You can still call difficulty and swim energy with current patch if you want
    ;; or just stop chasing this tick
    stop
  ]

  ;; 4. Determine how many patches to move this tick
  let steps-to-move ceiling travel-distance
  let move-patches sublist planned-path 0 (min list steps-to-move length planned-path)

  ;; Extra guard in case sublist yields empty list
  if empty? move-patches [
    stop
  ]

  ;; Store where the bass started this tick
  let old-patch patch-here
  
  ;; 5. Move along the planned patches in order
  foreach move-patches [ p ->
    
    ;; Safeguard: never face the same patch we are already on
    if p != patch-here [
      
      ;; Count patch visits
      if p != old-patch [
        ask p [
          set visits-by-stripedbass visits-by-stripedbass + 1
        ]
        set old-patch p
      ]
      
      ;; Move to the patch
      face p
      move-to p
      
      ;; Count time spent
      ask p [
        set ticks-spent-stripedbass ticks-spent-stripedbass + 1
      ]
    ]
  ]

  ;; 6. If we moved, compute difficulty of the final patch
  let next-step last move-patches
  ;calculate-chase-difficulty next-step
  
  ;; 7. Deduct hydrodynamic swimming energy
  calculate-swim-energy
  
  if not member? patch-here trail [
  set trail lput patch-here trail
]
end

;to chase-nearest-alewife ;; predator procedure
;    let alewife-prey prey-in-vision with [breed = alewives and (patchtype != "sea")] ; only chase alewives in outside of sea
;  if any? alewife-prey [
;    let nearest-alewife-prey min-one-of alewife-prey [distance myself]
;    if nearest-alewife-prey != nobody [
;      face nearest-alewife-prey
;    ]
;    
;    let travel-distance (speed / 3)
;;; 4. Determine how many patches to move this tick
;   fd travel-distance
;  ]
;end

to decide-chase-direction
  if [velocity] of patch-here > 0 [
    set chase-direction "seaward"
    calculate-difficulty-seaward   ;; uses vel inside logic
    calculate-swimming-speed-seaward
  ]
  if [velocity] of patch-here <= 0 [
    set chase-direction "landward"
    calculate-difficulty-landward
    calculate-swimming-speed-landward
  ]
end

to find-target-prey

  let candidates alewives with [
    [patchtype] of patch-here != "sea"
  ]

  ifelse any? candidates [
    ;; Filter for only edible prey (based on gape size)
    let edible-prey candidates with [
      body-depth <= [gape-size] of myself
    ]
    set target-prey min-one-of candidates [distance myself]
  ] [
    set target-prey nobody
  ]
end

to calculate-chase-difficulty [next-patch]

  let vel [velocity] of next-patch

  if chase-direction = "landward" [
    calculate-difficulty-landward   ;; uses vel inside logic
    calculate-swimming-speed-landward
  ]

  if chase-direction = "seaward" [
    calculate-difficulty-seaward
    calculate-swimming-speed-seaward
  ]
end

to calculate-cost-to-prey [prey-patch]

  ;; set initial conditions
  ask patch-here [ set cost-to-prey 1e9 ]
  ask prey-patch [ set cost-to-prey 0 ]

  ;; Start with a list containing the home patch
  let frontier (list prey-patch)
  let df difficulty-factor           ;; capture it locally

  while [not empty? frontier] [
    let current first frontier
    set frontier but-first frontier

    ask current [
      let current-cost cost-to-prey
      
      ask neighbors4 with [patch-terrain = "water"] [
        let travel-cost compute-travel-cost current self df
        let new-cost current-cost + travel-cost

        if new-cost < cost-to-prey [
          set cost-to-prey new-cost
          set frontier lput self frontier
        ]
      ]
    ]
  ]
  
end

to-report build-path-to-prey [start-patch prey-patch]
  let path (list start-patch)
  let cur start-patch
  let max-steps 500

  while [cur != prey-patch and max-steps > 0] [
    set max-steps max-steps - 1

    let candidates neighbors4 with [
      patch-terrain = "water" and
      cost-to-prey < [cost-to-prey] of cur
    ]

    if not any? candidates [ report path ]

    let next-p min-one-of candidates [cost-to-prey]
    set path lput next-p path
    set cur next-p
  ]

  report path
end
```

### Predator Speed

```{netlogo}
to adjust-stripedbass-speed ;; predator procedure
  let travel-distance (speed / 3)
  set prey-in-front alewives in-radius vision-distance 

  ifelse any? prey-in-front and energy > 0 ;; burst if there is a target and have energy
  [ set bursting? true ]
  [ set bursting? false ]

  ifelse bursting?
  [ if speed <= max-speed
    [ set speed (speed + (speed * 0.3))
      let burst-energy (E-swim * 0.3)
      set energy (energy - burst-energy) ]
  ]
  
  [ set speed (max-speed / 4) ]
end

```

### Consume Prey

```{netlogo}
to eat  ;; predator procedure
   if stomach-contents < stomach-capacity [
    ;; Find prey on the same patch
   let nearby (turtle-set alewives-here alewives-on neighbors)

    ;; Filter for only edible prey (based on gape size)
    let edible-prey nearby with [
      body-depth <= [gape-size] of myself
    ]

    if any? edible-prey [
      let meal one-of edible-prey
      ;; Digestible mass (75% of whole prey weight)
      let meal-weight ([weight] of meal) * 0.75

      ;; Remaining capacity
      let space-available (stomach-capacity - stomach-contents)

      ;; You can only eat as much as fits
      let intake (min (list meal-weight space-available))

      ;; Add biomass to stomach
      set stomach-contents (stomach-contents + intake)
      
      ;; add mercury
      ;; accumulates from prey eaten
      let mehg-prey-risk [mehg-total] of meal
      let hg-prey-risk [hg-total] of meal
      set mehg-foraging-undigested mehg-prey-risk
      set hg-foraging-undigested hg-prey-risk
      set stomach-contents-Hg (stomach-contents-Hg + hg-prey-risk);; mercury stomach amount
      set stomach-contents-MeHg (stomach-contents-MeHg + mehg-prey-risk);; methylmercury stomach amount
      
      ask meal [
        die
      ]
            ;; Update counters
      set completed-action? true
      set numAlewivesEaten numAlewivesEaten + 1
      set daytime-prey-eaten daytime-prey-eaten + 1
      mark-foraging-event
      
      ask patch-here [
      set prey-eaten-in-patch prey-eaten-in-patch + 1
      ]
    ]

  ]
 
end
```

